---
title: "Guided Topic Model (SpaTM-G)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guided_Topic_Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# SpaTM-G \|\| Guided Topic Model

This vignette looks at the Guided Topic Model (GTM) feature provided by SpaTM.

The model is based on the previously implemented GTM-decon method, now contained within an easy-to-use Rcpp-based package. The GTM extends the based topic model by allowing users to anchor topics to specific annotations from observations (cells or spots). For example, topics can be anchored to cell type annotations to ensure the learned topics represent cell types. This is practical cell type deconvolution where we can train the model on an annotated reference scRNA-seq dataset and apply it to a bulk dataset (or spot-based spatial data!). If a reference dataset is not available, we can train the model without a guided prior to see what transcriptional programs (topics) best represent the data.

```{r setup, message=FALSE,warning=F}
library(SpaTM)
library(SingleCellExperiment)
```

## Preparing a reference SingleCellTopicExperiment object

To start, we need a reference scRNA-seq dataset in the form of a SingleCellExperiment object. This can be created from a count matrix or converted from an existing Seurat or AnnData object. See the Seurat and zellkonverter packages for how to achieve both conversions.

Here we will simulate a dataset with 4 different cell types

```{r simulate_data}
set.seed(42)

# Parameters for simulation
n_genes <- 100
n_cells_per_type <- 50
n_cell_types <- 4
total_cells <- n_cells_per_type * n_cell_types

# Create base expression patterns for each cell type
base_patterns <- matrix(
  rgamma(n_genes * n_cell_types, shape = 2, scale = 1),
  nrow = n_genes,
  ncol = n_cell_types
)

# Create empty count matrix
counts <- matrix(0, nrow = n_genes, ncol = total_cells)
colnames(counts) <- paste0("cell_", 1:total_cells)
rownames(counts) <- paste0("gene_", 1:n_genes)

# Generate cell type labels
cell_types <- rep(paste0("celltype_", 1:n_cell_types), each = n_cells_per_type)

# Fill count matrix with simulated data
for (i in 1:n_cell_types) {
  start_idx <- ((i-1) * n_cells_per_type) + 1
  end_idx <- i * n_cells_per_type
  
  # Add noise and generate counts
  counts[, start_idx:end_idx] <- sapply(1:n_cells_per_type, function(x) {
    rpois(n_genes, lambda = base_patterns[, i] * runif(n_genes, 0.8, 1.2))
  })
}

# Create SingleCellExperiment object
sce <- SingleCellExperiment(
  assays = list(counts = counts),
  colData = DataFrame(cell_type = cell_types)
)

# Display dimensions and structure
sce
as.matrix(table(sce$cell_type))/ncol(sce)
```

With our SCE object, we can then convert it to a TopicExperiment, an object we created that inherits the SCE structure to facilitate integrating SpaTM analyses with other downstream tasks. During the initialization, you will either have to assign a number of topics to the `K` parameter (if running an unguided model) else you will need to set `guided = TRUE` and `labels = label` where `label` is the name of the column containing your cell annotations. These are expected to be a `factor` but will be converted if not.

```{r}
#Converting to a SingleCellTopicExperiment
sce <- SingleCellTopicExperiment(sce,
                                 guided = T,
                                 labels = 'cell_type')
K <- ncol(alphaPrior(sce))
D <- ncol(sce)
sce
```

Once the object is initialized, you can run the GTM model. The `num_threads` argument is made available to run the model with multiple threads to increase speed.

Here we can validate our training by seeing if the top topic weight for each cell matches its cell type annotation. In practice, this should be done on unseen samples and possibly include pseudo-bulked samples with known cell type proportions (see next code block).

```{r}
sce <- GTM(sce,K,D,num_threads = 1,maxiter = 50,
           verbal = F,zero_gamma = F,rand_gamma = F)

#Calculating parameter estimates
sce <- buildTheta(sce)
sce <- buildPhi(sce)

#Check training performance
sce$Pred <- levels(sce$cell_type)[apply(theta(sce),1,which.max)]
table(sce$cell_type,sce$Pred)
```

Topic inference on an unseen dataset is achieved with the `inferTopics` function. We provide the unseen dataset and the learned gene-by-topic matrix $\phi$ as input. From there, the learned observation-by-topics $\theta$ will represent our deconvolution results.

```{r}

## Creating pseudobulk of training dataset
bulk_mat <- as.matrix(rowSums(counts))
bulk_se <- SingleCellExperiment(assays = list(counts = bulk_mat))
bulk_se <- SingleCellTopicExperiment(bulk_se,
                                     K = K)
## Running Topic Inference
bulk_se <- inferTopics(bulk_se,maxiter = 100,phi = phi(sce),verbal = F)
bulk_se <- buildTheta(bulk_se)
colnames(theta(bulk_se)) <- colnames(phi(sce))

##Printing Deconvolution Results
theta(bulk_se)
```

```{r}
sessionInfo()
```
